The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Nature or tidal nurture?

### 2025-09-29 thesis/discussion.tex
thesis edits

Closely related to our work, \citet{iorio+2019} applied idealized N-body
simulations to study tidal effects on Scl. They similarly found weak
tidal effects, even for a dark-matter-free model. However, their orbits
are less well-constrained and they did not consider the LMC.

@@

@@

@@

As a tidal origin of Scl and UMi's extended density profiles is
disfavoured, we consider alternative processes which may form a dwarf
galaxy ``stellar halo''. Some recent works have also confirmed the
presence of extended, likely non-tidal, density profiles. For example,
\citet{chiti+2021}; \citet{chiti+2023} spectroscopically confirm members
out to \(~9 R_h\) in Tucana II. Given these stars are misaligned with
the orbit and the lack of a velocity gradient, tides seem to be an
unlikely explanation for Tucana II as well. The prevalence of stellar
halos around dwarf galaxies remains an active research topic.

Many dwarf galaxies also host multiple chemodynamical stellar
populations. Typically, older populations are extended, kinematically
hotter, and metal-poo , whereas the younger populations are more
compact, colder, and metal-rich. Examples include Carina
\citep[\citet{fabrizio+2016}, \citet{kordopatis+2016}]{battaglia+2012},
Fornax \citep[\citet{amorisco+evans2012},
\citet{delpino+aparicio+hidalgo2015}]{battaglia+2006}, Sextans
\citep{battaglia+2011, cicuendez+battaglia2018, roederer+2023}, and
Andromeda II
\citep{mcconnachie+arimoto+irwin2007, ho+2012, delpino+2017}. Evidence
of multiple-populations in dwarf galaxies suggests that effects like
galaxy mergers or episodic star formation shape the stellar structure of
dwarf galaxies.

@@

\textbf{Dynamical heating of old stars}. In a galaxy, older stars
generally have higher random velocities (i.e., kinematically ``hotter'')
than younger stars. In dwarf galaxies, several mechanism may heat
stellar components, including supernovae kicks in forming stars,
feedback-driven potential fluctuations
\citep{stinson+2009, maxwell+2012, el-badry+2016, mercado+2021}, or
heating by dark sub-subhalos \citep{penarrubia+2025}. Over time, these
processes naturally produce a more extended, older stellar population.

\textbf{Episodic star formation and feedback.} Dwarf galaxies are
thought to experience bursty star formation--i.e., consisting of several
episodes of intense star formation separated by periods of quiescence
\citep[e.g.,][]{salvadori+ferrara+schneider2008, valcke+derijcke+dejonghe2008, wheeler+2019, azartash-namin+2024}.
Stellar feedback in a dwarf galaxy's shallow potential well drives
oscillations in the star formation rate. Alternatively, re-ionization
may temporarily suspending star formation \citep{benitez-llambay+2015}.
In many simulations, shrinking gas reservoirs form
centrally-concentrated later generations of stars, naturally spawning
multiple populations \citep{kawata+2006, revaz+jablonka2018}. However,
star formation is also sensitive to a dwarf's environment.

@@

\textbf{Major mergers.} When galaxies merge, they may leave signatures
such as population gradients and stellar halos. Classical dwarfs have a
\(\sim 10\%\) chance of undergoing a major merger since redshift \(z=1\)
\citep{deason+wetzel+garrison-kimmel2014}. Mergers may preferentially
disperse the lower-mass galaxy's stars, forming a halo and a metallicity
gradient \citep{benitez-llambay+2016}. Intermediate-mass mergers (with
mass-ratios of \(\sim\) 1:5) most effectively populate halos, balancing
stellar mass from larger galaxies with better dispersion of lower-mass
galaxies \citep{deason+2022}. Tuc II's properties are suggested to
originate from a similar merger
\citep{tarumi+yoshida+frebel2021, querci+2025}.

\textbf{Gas-rich mergers.} If a merger occurs between gas-rich galaxies,
triggered star formation may occur in the aftermath
\citep[e.g.,][]{genina+2019}. And II and Phoenix have steep metallicity
gradients and unusual prolate rotation, theorized to result from mergers
of disky dwarfs \citep{lokas+2014, fouquet+2017, cardona-barrero+2021}.

\textbf{Minor mergers / accretion}. Like how the Milky Way's halo is
believed to be built from many minor mergers, dwarf galaxies may form
halos through accretion of yet fainter dwarfs.
\citet{ricotti+polisensky+cleland2022} demonstrated that accretion of
ultra-faint dwarfs may create dwarf stellar halos. A stream detected
around And II further supports the occurrence of mergers among dwarfs
\citep{amorisco+evans+vandeven2014, roederer+2023}. In this scenario,
the halo chemistry would resemble a population of ultra-faint dwarfs.
The occurance of dwarf mergers would further constrain small-scale
galaxy formation.

\textbf{Tidal preprocessing}. Dwarf galaxies may have been
``preprocessed'' by a larger satellite like the LMC
\citep[e.g.,][]{santistevan+2023, riley+2024}. From the orbit
integrations above, it is possible that UMi was once an LMC satellite.
However, the prevalence of preprocessing remains uncertain. Like stellar
heating, preprocessing redistributes already-present stellar
populations.

@@

\textbf{Modified Newtonian Dynamics (MOND)}. MOND modifies gravity
instead of using dark matter to explain the rotation curves of galaxies.
In MOND, tides may more strongly affect dwarf galaxies owing to the lack
of dark matter mass and the stronger MW tidal field
\citep{mcgaugh+wolf2010, brada+milgrom2000}. A tidal origin of the
density excess is more likely in this case. If MOND is to recover the
observed velocity dispersions of satellites, then many more dwarfs may
be actively tidally disrupting as well
\citep{casas+2012, yang+2014, wang+2024a}.

@@

\textbf{Chemistry}. Large samples of detailed chemical abundances have
been invaluable for understanding MW substructure (e.g.~for
\emph{Gaia}-Sausage Enceladus). Chemistry, particularly comparing the
inner and outer regions could test if the halo appears to originate from
a distinct system than the dwarf, differentiating many \emph{internal}
versus \emph{external} scenarios.

\textbf{Star formation histories}. Evidence of significant star
formation episodes or lack thereof may differentiate scenarios which
rely on strong star formation bursts (e.g.~\emph{episodic star formation
history, induced star formation, and gas-rich mergers.})

\textbf{Kinematics.} Particularly, for models relying on recent tidal
disruption, kinematic disequilibrium features should be visible. These
would appear as velocity gradients, increasing velocity dispersions,
outward-biased moving stars, or non-phase mixed structures
\citep[e.g,][]{kroupa1997, read+2006, sanchez-salcedo+hernandez2007}.

\textbf{Deep photometry} may find or rule out signs of dynamical
disequilibrium and tidal tails for tidally susceptible models
\emph{(e.g., MOND and tidal dwarfs.)}

>>>>

@@

Our work is inconsistent with claims of observational tidal disruption.
Closely related to our work, \citet{iorio+2019} applied idealized N-body
simulations to study tidal effects on Scl. They similarly found weak
tidal effects, even for a dark-matter-free model. However, their orbits
are less well-constrained and they did not consider the LMC. If the
extended density profile of Scl is formed through another mechanism
which also creates multiple populations (see discussion below), then
both observational features would become consistent with such an
explanation.

@@

As our models disagree with a tidal interpretation for the density
profile of Ursa Minor, the observed density excess likely originate from
another process. However, also find no evidence for a velocity gradient
with current spectroscopic samples Section~\ref{sec:extra_rv_models}.
While the origin of inner substructure in UMi is beyond the scope of
this work, this structure, if real, points towards a complex origin of
Ursa Minor.

@@

As we disfavour a tidal origin of Scl and UMi's extended density
profiles, we consider alternative processes which may form a dwarf
galaxy ``stellar halo''. Some recent works have also confirmed the
presence of extended, likely non-tidal, density profiles. Tucana II is
another dwarf with spectroscopic confirmation of outer ``halo'' stars
out to \(9R_h\) and inconsistent with a tidal origin
\citep{chiti+2021, chiti+2023}. The prevalence of stellar halos around
dwarf galaxies remains an active research topic.

Many dwarf galaxies also host multiple chemodynamical stellar
populations. Typically, older populations are extended, kinematically
hotter, and metal-poo, whereas the younger populations are more compact,
colder, and metal-rich. Examples include Carina
\citep[\citet{fabrizio+2016}, \citet{kordopatis+2016}]{battaglia+2012},
Fornax \citep[\citet{amorisco+evans2012},
\citet{delpino+aparicio+hidalgo2015}]{battaglia+2006}, Sextans
\citep{battaglia+2011, cicuendez+battaglia2018, roederer+2023}, and
Andromeda II
\citep{mcconnachie+arimoto+irwin2007, ho+2012, delpino+2017}. Evidence
of multiple-populations in dwarf galaxies suggests that effects like
galaxy mergers or episodic star formation shape the stellar structure of
dwarf galaxies.

@@

\textbf{Dynamical heating of old stars}. In a galaxy, older stars
generally have higher random velocities (i.e., kinematically ``hotter'')
than younger stars. In dwarf galaxies, several mechanism may heat
stellar components, including supernovae kicks in forming stars,
feedback-driven potential fluctuations
\citep{stinson+2009, maxwell+2012, el-badry+2016, mercado+2021}, or
heating by dark sub-subhalos \citep{penarrubia+2025}. Over time, these
processes naturally produce a more extended, older stellar population.
If internal dynamical heating explains stellar halos, then similar
stellar halos should exist around similar dwarfs.

\textbf{Episodic star formation and feedback.} Dwarf galaxies are
thought to experience bursty star formation--i.e., consisting of several
episodes of intense star formation separated by periods of quiescence
\citep[e.g.,][]{salvadori+ferrara+schneider2008, valcke+derijcke+dejonghe2008, wheeler+2019, azartash-namin+2024}.
Stellar feedback in a dwarf galaxy's shallow potential well drives
oscillations in the star formation rate. Alternatively, re-ionization
may temporarily suspending star formation \citep{benitez-llambay+2015}.
In many simulations, shrinking gas reservoirs form
centrally-concentrated later generations of stars, naturally spawning
multiple populations \citep{kawata+2006, revaz+jablonka2018}. If
episodic or vigorous star formation helps create stellar halos, then
corresponding evidence of bursty star formation histories should be
apparent.

@@

\textbf{Major mergers.} When galaxies merge, they may leave signatures
such as population gradients and stellar halos. Classical dwarfs have a
\(\sim 10\%\) chance of undergoing a major merger since redshift \(z=1\)
\citep{deason+wetzel+garrison-kimmel2014}. Mergers may preferentially
disperse the lower-mass galaxy's stars, forming a halo and a metallicity
gradient \citep{benitez-llambay+2016}. Intermediate-mass mergers (with
mass-ratios of \(\sim\) 1:5) most effectively populate halos, balancing
stellar mass from larger galaxies with better dispersion of lower-mass
galaxies \citep{deason+2022}. Tuc II's properties are suggested to
originate from a similar merger
\citep{tarumi+yoshida+frebel2021, querci+2025}. Major mergers should
produce multiple chemodynamical populations consistent with two
distinct, but similar mass dwarf galaxies.

\textbf{Gas-rich mergers.} If a merger occurs between gas-rich galaxies,
triggered star formation may occur in the aftermath
\citep[e.g.,][]{genina+2019}. And II and Phoenix have steep metallicity
gradients and unusual prolate rotation, theorized to result from mergers
of disky dwarfs \citep{lokas+2014, fouquet+2017, cardona-barrero+2021}.
Gas-rich mergers should have at least 3-distinct chemodynamical
populations, one from each galaxy and one from the induced starburst.

\textbf{Minor mergers / accretion}. Like how the Milky Way's halo is
believed to be built from many minor mergers, dwarf galaxies may form
halos through accretion of yet fainter dwarfs.
\citet{ricotti+polisensky+cleland2022} demonstrated that accretion of
ultra-faint dwarfs may create dwarf stellar halos. A stream detected
around And II further supports the existence of mergers among dwarfs
\citep{amorisco+evans+vandeven2014, roederer+2023}. The occurrence of
dwarf mergers would further constrain small-scale galaxy formation. In
this scenario, the halo chemistry would resemble a population of
ultra-faint dwarfs.

\textbf{Tidal preprocessing}. Dwarf galaxies may have been
``preprocessed'' by a larger satellite like the LMC
\citep[e.g.,][]{santistevan+2023, riley+2024}. From the orbit
integrations above, it is possible that UMi was once an LMC satellite.
However, the prevalence of preprocessing remains uncertain. Like stellar
heating, preprocessing redistributes already-present stellar populations
without creating new populations, but would occur more stochastically.
Distant, unbound stream stars may be found around a preprocessed dwarf.

@@

\textbf{Modified Newtonian Dynamics (MOND)}. MOND modifies gravity
instead of using dark matter to explain the rotation curves of galaxies.
In MOND, tides may more strongly affect dwarf galaxies owing to the lack
of dark matter mass and the stronger MW tidal field
\citep{mcgaugh+wolf2010, brada+milgrom2000}. A tidal origin of the
density excess is more likely in this case. If MOND is to recover the
observed velocity dispersions of satellites, then many more dwarfs may
be actively tidally disrupting
\citetext{\citealp{mcgaugh+wolf2010}; \citealp[but see
also][]{sanchez-salcedo+hernandez2007}}.

@@

\emph{Chemistry}. Large samples of detailed chemical abundances have
been invaluable for understanding MW substructure (e.g.~for
\emph{Gaia}-Sausage Enceladus). Chemistry, particularly comparing the
inner and outer regions could test if the halo appears to originate from
a distinct system than the dwarf, differentiating many \emph{internal}
versus \emph{external} scenarios.

\emph{Star formation histories}. Evidence of significant star formation
episodes or lack thereof may differentiate scenarios which rely on
strong star formation bursts (e.g.~\emph{episodic star formation
history, induced star formation, and gas-rich mergers.})

\emph{Kinematics}. Particularly, for models relying on recent tidal
disruption, kinematic disequilibrium features should be visible. These
would appear as velocity gradients, increasing velocity dispersions,
outward-biased moving stars, or non-phase mixed structures
\citep[e.g,][]{kroupa1997, read+2006, sanchez-salcedo+hernandez2007}.

\emph{Deep photometry} may find or rule out signs of dynamical
disequilibrium and tidal tails for tidally susceptible models
\emph{(e.g., MOND and tidal dwarfs.)} In addition, photometry will help
constrain the prevalence and nature of extended features in dwarf
galaxies.

### 2025-09-29 thesis/extra_results.tex
thesis edits

In this section, we briefly explore additional simulations testing
assumptions not varied in the main text: variations in the halo
concentration, cored halos, alternative orbits, anisotropy, and
oblateness.

@@

>>>>

In this section, we briefly explore additional simulations testing
assumptions not varied in the main text: variations in the halo
concentration, cored halos, alternative orbits, anisotropy, and
ellipsoidal models.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/umi_sim_stream.pdf}}
\caption[Ursa Minor predicted stream]{The properties of the stream
around the UMi \smallperi{} orbit with Plummer stars. The panels are all
as a function of \(\xi'\), the distance along the stream as defined by
the current GSR proper motion vector. The top panels show the GSR proper
motions in RA and Dec, and the bottom two show the distance and GSR
radial velocities. To sample the stream, we randomly draw 100,000
samples from the snapshot based on the stellar weights. A detectible
gradient in \(\mu_{\alpha*}\) and LOS velocity should be detectible if
the stream is tracked across several
degrees.}\label{fig:umi_tidal_stream}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_sim_stream.pdf}}
\caption[Scl predicted stream]{The predicted properties of a tidal tail
in the Scl model.}\label{fig:scl_tidal_stream}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_mw_impact_stream.pdf}}
\caption[Scl predicted stream]{the predicted properties of the stream of
stars around Scl in the MW impact model. Note that, compared to
Fig.~\ref{fig:scl_sim_stream}, this model produces a much more scattered
and misaligned stream, due to the long time since stream formation and
nonlinear effects of the LMC encounter. \textbf{todo check 2d velocity
gradients in this case}}\label{fig:scl_mw_impact_tidal_stream}
\end{figure}
